---
title: "HW8"
author: "Guannan Shen"
date: "November 19, 2018"
output: 
  pdf_document:
    latex_engine: lualatex
    number_sections: yes
    toc: yes
    toc_depth: 5
  word_document:
    toc: yes
    toc_depth: '5'
  html_document:
    number_sections: yes
    theme: united
    toc: yes
    toc_depth: 5
    toc_float: yes
---

```{r setup, include=FALSE, cache = FALSE}
require("knitr")
opts_chunk$set(tidy.opts=list(width.cutoff=60),tidy=TRUE)
opts_chunk$set(engine = "R")
knitr::opts_chunk$set(echo = T)
knitr::opts_chunk$set(message = F)
knitr::opts_chunk$set(warning = F)
## setting wd in DELL
## opts_knit$set(root.dir = "~/Stats/CIDA_OMICs/CIDA_OMICS/7659Stats_Genetics/HW5/")
## setting working directory in asus 
## opts_knit$set(root.dir = "C:/Users/hithr/Documents/Stats/CIDA_OMICs/7659Stats_Genetics/HW5/") 
## setting working directory in ubuntu
opts_knit$set(root.dir = "~/Documents/Stats/CIDA_OMICs/7659Stats_Genetics/HW8/")
                                                 
## cache = F, if cache = T, will not revaluate code chunk everytime
## double or more space to insert a line break
```


```{r libs}
## set up workspace

library(bPeaks)
library(knitr)
library(tidyverse)

library(magrittr)
library(FlowSorted.Blood.450k)
library(quadprog)
library(minfi)
library(IlluminaHumanMethylation450kmanifest)
library(stats)
options(stringsAsFactors = F)
options(dplyr.width = Inf)
getwd()
## not in function
'%nin%' <- Negate('%in%')

# ######## clean memory ######################
# rm(list = ls())
# gc()

```

# 1. Cell Type Composition
Background on data set: The occurrence of gestational diabetes mellitus (GDM) during pregnancy is believed to alter obesity risk of offspring later in life. It was hypothesized that exposure to maternal GDM in utero will be associated with changes in DNA methylation patterns of key genes and pathways in the offspring, which will mediate the association between in utero exposure and childhood adiposity-related outcomes. Data was collected from the EPOCH (Exploring Perinatal Outcomes in CHildren) study, a historical prospective cohort that enrolled children aged 10.5 on average (T1) who were exposed or not exposed to maternal GDM during the intrauterine life. DNA was extracted from peripheral blood samples collected from children at the T1 EPOCH visit on 85 exposed to GDM and 85 unexposed to GDM, and methylation data was generated using the Illumina Infinium HumanMethylation450 BeadChip. We have provided data on a subset of 10 subjects (all Non-Hispanic Whites) from these two groups to identify methylation sites and nearby genes that show differential methylation between the two groups. The data for this problem is available through a link on Canvas to the "blood" folder on Dropbox.

## (a) Use SWAN normalization from Homework 7, then find differentially methylated positions based on exposure status using dmpFinder(). Are there any DMPs with q-value <= 0.10 (or p-value cutoff of 10-5)? Summarize the results in a table and include the direction (hyper or hypo methylated based on exposure status).

```{r SWANDMPS}
## import data 
## under directory 
## ~/Documents/Stats/CIDA_OMICs/7659Stats_Genetics/HW8/
baseDir1 <- "blood/plate1"
targets1 <- read.metharray.sheet(baseDir1)
baseDir2 <- "blood/plate2"
targets2 <- read.metharray.sheet(baseDir2)
targets <- rbind(targets1, targets2)

rgSet <- read.metharray.exp(targets=targets, extended=T)
#for part c), use "extended =F"
rgSet[[1]]
# clinical and groups
sampleNames(rgSet) = rgSet[[1]]
getManifest(rgSet)
clindat <- read.table("blood/demographic.txt", sep="\t", header=T)
## grouping 
class(clindat$Exposure)
clindat$child_sex
pData(rgSet)$Sample_Group <- clindat$Exposure
pData(rgSet)$child_sex <- clindat$child_sex

# swan normalization and exposure groups
mset <- preprocessRaw(rgSet)
msetSWAN <- preprocessSWAN(rgSet)
# m values 
mvalueset <- getM(msetSWAN)
## dmp by exposure
dmp_set <- dmpFinder(mvalueset, pheno = clindat$Exposure, type = c("categorical") )

#sig table
# 0 rows
dmp_set[dmp_set$qval <= 0.1, ]
# 4 rows
dmp_settop <- dmp_set[dmp_set$pval <= 0.00001, ] %>% 
  mutate(methy_status = if_else(.$intercept > 0, "hyper-methylated", "hypo-methylated"),
         CpGs = rownames(.)) %>% dplyr::select(CpGs, everything())

kable(dmp_settop, caption = "Top CpGs by Exposure Status", align = "c")

```

The direction of methylation was defined by the value of intercept, positive intercept means hyper-methylation in exposure status 1. 



## (b) Because blood is a heterogeneous collection of different cell types, it has been recognized that it may be important to adjust for cell-type composition in your analysis. Why is cell-type composition relevant for DNA methylation studies? What are the methods available to adjust for cell-type composition when celltypes are not directly measured (as in this example)?

Blood is a heterogeneous collection of different cell types, each with a very different DNA methylation profile. Basically, DNA methylation has a very high variability on cell type level. In the Jaffe & Irizarry, Genome Biology (2014) paper, they showed in five previously published studies, cellular composition explained much of the observed variability in DNA methylation. That is, cellular composition is a confounder in blood sample methylation study.  

Methods to estimate cell counts have been developed by different approaches, such as reference-free deconvolution, reference-based deconvolution and Bayesian semi-supervised method (BayesCCE). In this example, the Houseman reference-based method, estimateCellCounts(), implemented in the R package minfi is the optimal method for reference-based deconvolution.

## (c) Estimate cell counts using estimateCellCounts(). Explain the graph that is displayed by this function. Then repeat part a), but include cell-type composition as covariates in your model. Summarize your results again in a table. How do the results compare between the unadjusted and adjusted cell-type composition analysis? Provide possible explanations if you do not see differences.Based on the available clinical data file, discuss what other covariates that you may consider including in the model (no need to implement).

```{r celltype}
rgSet_F <- read.metharray.exp(targets=targets, extended=F)
sampleNames(rgSet_F) = rgSet_F[[1]]
pData(rgSet_F)$Sample_Group <- clindat$Exposure
pData(rgSet_F)$child_sex <- clindat$child_sex

# swan normalization and exposure groups
mSet_F <- preprocessRaw(rgSet_F)
mSet_FSWAN <- preprocessSWAN(rgSet_F)
# m values 
mvalueSet_F <- getM(mSet_FSWAN)
# estimate cell
cellcount <- estimateCellCounts(rgSet_F, compositeCellType = "Blood", meanPlot = T, 
                                cellTypes = c("CD8T","CD4T", "NK","Bcell","Mono","Gran"))
# samples and groups 
sum(rownames(cellcount) != rgSet_F[[1]])
cell <- as.data.frame(cellcount)
cell$exposure <- clindat$Exposure
# have a look
cell[1,]

# reference with dmpfinder
## dmp by exposure
dmp_Set_F <- dmpFinder(mvalueSet_F, pheno = clindat$Exposure, type = c("categorical") )
dmp_Set_Ftop <- dmp_Set_F[dmp_Set_F$pval <= 0.00001, ] %>% 
  mutate(methy_status = if_else(.$intercept > 0, "hyper-methylated", "hypo-methylated"),
         CpGs = rownames(.)) %>% dplyr::select(CpGs, everything())

kable(dmp_Set_Ftop, caption = "Top CpGs by Exposure Status", align = "c")

# lm 
n_cpgs <- nrow(mvalueSet_F)
outcome_lm <-  lapply(1:n_cpgs, function(i){
  lm = lm(mvalueSet_F[i,] ~ exposure + Bcell + CD4T +
            CD8T + Gran + Mono + NK, data = cell)
  coef = c(summary(lm)$coefficients[2, 1], 
           summary(lm)$coefficients[2, 4])
  if(i %in% seq(0, 500000, by = 20000)){
    print(i)
  }
  return(coef)
  
})
outcome_lm <-  data.frame(matrix(unlist(outcome_lm), ncol = 2, byrow = TRUE,
                          dimnames = list(
                     c(rownames(mvalueSet_F)), c("Estimate", "p.value"))))

outcome_lm <- outcome_lm %>% mutate(CpGs = rownames(mvalueSet_F)) %>% dplyr::select(CpGs, everything())
outcome_lm <- outcome_lm[order(outcome_lm$p.value, decreasing = F), ]

outcome_lm1 <- outcome_lm[outcome_lm$p.value <= 0.00001, ] %>% 
  mutate(methy_status = if_else(.$Estimate > 0, "hyper-methylated", "hypo-methylated")) %>% 
  dplyr::select(CpGs, everything())

kable(outcome_lm1, caption = "Top CpGs by Exposure Status, adjusting for cell types", align = "c")

 ## compare 
cpg_com <- dmp_Set_Ftop$CpGs[dmp_Set_Ftop$CpGs %in% outcome_lm1$CpGs]

```


The meanPlot is the average DNA methylation across the cell-type discrimating probes within the mixed and sorted samples. The table above showed the CpGs with p values less than 0.00001. The results between the unadjusted and adjusted cell-type composition analysis differ a lot. Only one CpG `r cpg_com` was shared in both approaches. The whole linear regression has the M-values of CpGs as the outcome (which is bimodal, might apply log10 transformation), and exposure, cell types as covariates. The direction here was defined by the value of the beta estimates of the exposure status. Both the directions and scopes of exposure effects on CpGs M values change a lot. 

The gender of child may also be included in the model. 

# 2. ChIP-Seq
• Download the data provided on Canvas (tup1 IP.txt, mock IP.txt, input IP.txt). These are three ChIP-Seq experiments in yeast from Park et al. 2013 PLoS One 8:12 e83506 (http://www.ncbi.nlm.nih.gov/pubmed/24349523). Tup1 is a transcriptional repressor and the mock and input are two different controls for comparison.

## (a) By examining the GEO links and reference, what methods were used for sequencing, basecalling, mapping reads and dealing with non-uniquely mapped reads? What is the difference between mock and input controls?

Sequencing was carried out by either Illumina HiSeq 2000 or SOLiD V4. Basecalls performed using CASAVA version 1.8. ChIP-seq reads were aligned to the sacCer3 genome assembly using BWA (Version: 0.5.9-r16) with default options. Non-uniquely mapped reads were filtered out in order to remove the reads with low mapping quality.  

Mock ChIP DNA was prepared by immunoprecipitation with IgG Sepharose in the wild type strain with no TAP-tagged protein expression. Input DNA was prepared in parallel with the SWI6-TAP ChIP sample but leaving out the immunoprecipitation step. So the mock had the no-antibody ip step, but the input did not have the ip step.

## (b) Using baseLineCalc(), what is the average number of sequenced mapped at each position? Does the Tup1 ChIP or mock sample have more average reads? (Hint: This function and the function in part c) only need the last column of read counts from allData$IPdata and allData$controlData)

```{r base}
data(yeastCDS) #for gene location annotation
## also input_IP.txt
allData <- dataReading("tup1_IP.txt", "mock_IP.txt", yeastSpecies = yeastCDS$Saccharomyces.cerevisiae) 
#read in data

## average test group
mean_tup1 <- baseLineCalc(allData$IPdata[,3])

## average control group
mean_mock <- baseLineCalc(allData$controlData[,3])

##
mean_tup1 > mean_mock

```

The average number of sequenced mapped at each position for Tup1 ChIP is `r mean_tup1`, for mock control is `r mean_mock`. The Tup1 ChIP has more average reads. 

## (c) Examining only chromosome V ("chrV"), by subsetting the first column of allData$IPdata and allData$controlData, run peakDetection(), with the baseLineIP and baseLineControl values calculated in part b).

```{r peak}
## chromosomes names
chromNames <- unique(allData$IPdata[,1])
chromNames[5]
## mask
filter <- stats::filter
## 
IPsignal <-  dataSmoothing(allData$IPdata[allData$IPdata[,1] == "chrV",3], 20)
controlSignal <- dataSmoothing(allData$controlData[allData$controlData[,1] == "chrV",3], 20)
# peak detection
detectedPeaks <-  peakDetection(IPdata = IPsignal, controlData = controlSignal, chrName = as.character(chromNames[5]), windowSize = 150, windowOverlap = 50, outputName = paste("bPeaks_example_", chromNames[5], sep = ""), baseLineIP = mean_tup1, baseLineControl = mean_mock, IPthreshold = 6, controlThreshold = 4, ratioThreshold = 2, averageThreshold = 0.7, peakDrawing = TRUE)
# print detected genomic positions
print(detectedPeaks)

# 
nrow(detectedPeaks)

## peak locations
peakLocation(bedFile = "bPeaks_example_chrV.bed", cdsPositions = yeastCDS$Saccharomyces.cerevisiae, withoutOverlap = FALSE, outputName = "bPeaks_example_chrV", promSize = 800)

## import the result
promo <- read.delim("bPeaks_example_chrV_peakLocation_inPromoters.txt", header = F, sep = "\t")
promos <- NULL
for(i in 1:nrow(promo)){
promos[i] <- unlist(strsplit(unlist(strsplit(promo[i,3], "|", fixed=TRUE))[3], "=", fixed = T))[2]
}
promos
cds <- read.delim("bPeaks_example_chrV_peakLocation_inCDS.txt", header = F, sep = "\t")
cdss <- NULL
for(i in 1:nrow(cds)){
cdss[i] <- unlist(strsplit(unlist(strsplit(cds[i,3], "|", fixed=TRUE))[3], "=", fixed = T))[2]
}
cdss

```

This method detected `r nrow(detectedPeaks)` peaks. This method using a sliding window to scan the genomic sequence. Four criterion are used to define the peak region, peak shape, and adjacent regions may be merged. Basically, to find the significant window first and then right peak shapes within the window. 

These 4 parameters are used to define the interesting region. T1 = IPthreshold is a high number of reads in the IP sample. T2 = controlThreshold is a low number of reads in the control sample. "ratioThreshold" is a high value of log(IP/control) and "averageThreshold" is a good sequencing coverage for both IP and control samples.  

"peak_datasummary" shows graphical representations of the detected basic peaks, together with the values of the parameters used to detect the region, and the number of windows and peaks.  


"bPeaks-Drawing" 1st page showed the IP and control peak shape of peak #1 versus the location of this peak. Basically, by just plotting the 4 parameters of this peak against the location on the chromosome of this peak. 

Find the attached pdf at the end of the document, "example" using the mock and "input" using the input as control. 

## (d) The bPeaks package uses simple fold change cutoffs. What alternative methods discussed in class would you apply for a more rigorous approach to detect peaks using a statistical testing framework? Describe the methods and statistical approach.


In the class, we discussed using empirical FDR or Irreproducible Discovery Rate (IDR). The IDR using the pairs of replicates, ranking the peaks by reproducibility, and giving each peak an index. At the end, overall expected rate of irreproducible discoveries (IDR) is calculated. 

## (e) Using peakLocation() with the file provided in Canvas "bPeaks results.bed" and yeastCDS$Saccharomyces.cerevisiae for the cdsPositions. How many of the peaks are in genes or promoters? What are those genes?

There are 15 peaks in promoters and 2 peaks in CDS (coding region of the gene). Promoters are `r promos`. And cds are `r cdss`. 

## f) Now repeat b), c) and e) using the input IP sample instead ("input IP.txt"). What differences do you find between the results using mock or input IP? How does that relate to the conclusions in the paper (see Discussion)?

```{r input}

## also input_IP.txt
allData <- dataReading("tup1_IP.txt", "input_IP.txt", yeastSpecies = yeastCDS$Saccharomyces.cerevisiae)

## average test group
mean_tup1 <- baseLineCalc(allData$IPdata[,3])

## average control group
mean_input<- baseLineCalc(allData$controlData[,3])
##
mean_tup1 > mean_input

## chromosomes names
chromNames <- unique(allData$IPdata[,1])
chromNames[5]
## mask
filter <- stats::filter
## 
IPsignal <-  dataSmoothing(allData$IPdata[allData$IPdata[,1] == "chrV",3], 20)
controlSignal <- dataSmoothing(allData$controlData[allData$controlData[,1] == "chrV",3], 20)

# peak detection
detectedPeaks <-  peakDetection(IPdata = IPsignal, controlData = controlSignal, chrName = as.character(chromNames[5]), windowSize = 150, windowOverlap = 50, outputName = paste("bPeaks_input_", chromNames[5], sep = ""), baseLineIP = mean_tup1, baseLineControl = mean_input, IPthreshold = 6, controlThreshold = 4, ratioThreshold = 2, averageThreshold = 0.7, peakDrawing = TRUE)
# print detected genomic positions
print(detectedPeaks)

# 
nrow(detectedPeaks)

## peak locations
peakLocation(bedFile = "bPeaks_input_chrV.bed", cdsPositions = yeastCDS$Saccharomyces.cerevisiae, withoutOverlap = FALSE, outputName = "bPeaks_input_chrV", promSize = 800)

## import the result
promo_in <- read.delim("bPeaks_input_chrV_peakLocation_inPromoters.txt", header = F, sep = "\t")
promos_in <- NULL
for(i in 1:nrow(promo_in)){
promos_in[i] <- unlist(strsplit(unlist(strsplit(promo_in[i,3], "|", fixed=TRUE))[3], "=", fixed = T))[2]
}
promos_in
cds_in <- read.delim("bPeaks_input_chrV_peakLocation_inCDS.txt", header = F, sep = "\t")
cdss_in <- NULL
for(i in 1:nrow(cds_in)){
cdss_in[i] <- unlist(strsplit(unlist(strsplit(cds_in[i,3], "|", fixed=TRUE))[3], "=", fixed = T))[2]
}
cdss_in
## difference 
 sum(promos_in %in% promos == 0)

```


With input_IP as control, `r nrow(detectedPeaks)` peaks were detected. 18 peaks are in the promoter region and 2 peaks are in the CDS region. Promoters are `r promos_in`. And cds are `r cdss_in`. I found the genes of CDS were identical by these two approaches. And by using input DNA as the control, `r sum(promos_in %in% promos == 0)` more genes (promoter region) were detected, others were shared by these 2 different control approaches. 

In the paper discussion, they found there was a background signal deriving from expression bias by input DNA control. Thus, the background signal of transcription factor binding cannot be canceled. By using the corresponding mock ChIP data as the control, one can minimize false positives.

According to the discussion, the `r sum(promos_in %in% promos == 0)` more genes (promoter region) found by input DNA control might be false positive, might due to the expression bias. 


