---
title: "7659 HW3"
author: "Guannan Shen"
date: "October 4, 2018"
output: 
  word_document:
    toc: yes
    toc_depth: '5'
  pdf_document:
    latex_engine: lualatex
    number_sections: yes
    toc: yes
    toc_depth: 5
  html_document:
    number_sections: yes
    theme: united
    toc: yes
    toc_depth: 5
    toc_float: yes
---


```{r setup, include=FALSE, cache = FALSE}
require("knitr")
opts_chunk$set(tidy.opts=list(width.cutoff=60),tidy=TRUE)
opts_chunk$set(engine = "R")
knitr::opts_chunk$set(echo = T)
knitr::opts_chunk$set(message = F)
knitr::opts_chunk$set(warning = F)
## setting working directory in asus 
opts_knit$set(root.dir = "C:/Users/hithr/Documents/Stats/CIDA_OMICs/7659Stats_Genetics/HW3/")
## setting working directory in ubuntu
## opts_knit$set(root.dir = "~/Documents/Stats/gitlab/ECCHO_Guannan/Code/")
                                                 
## cache = F, if cache = T, will not revaluate code chunk everytime
## double or more space to insert a line break
```


```{r libs}
## set up workspace
library(knitr)
library(tidyverse)
library(samr)
library(impute)
library(limma)
library(gtools)
library(qvalue)
options(stringsAsFactors = F)
options(dplyr.width = Inf)
getwd()
```

# T-statistics of microarray data
## 1. For each gene, calculate the fold change between the knock-out and wildtype groups. List the top 10 genes that show the largest fold change (positive or negative). 
The fold change is calculated by knock-out over wildtype groups. 

```{r ttest1}
# read in raw data 
apodata <- read.table("hw3arraydata.txt", header = TRUE)
dim(apodata)
aponame <- read.table("hw3genenames.txt", header = FALSE, blank.lines.skip = FALSE)
dim(aponame)
colnames(aponame) <- "genenames"

# combine the names and intensity
ai <- cbind(aponame, apodata)


## the raw data is log2 transformed 
## the fold change is calculated from the subtraction
ailogratio <- (base::rowSums(ai[,10:17]) - base::rowSums(ai[,2:9]))/8
aifc <- logratio2foldchange(ailogratio)
aifc <- data.frame(ai$genenames, aifc)

# test missing data
kable(apply(ai[, 2:17] ,2, function(x){sum(is.na(x))}), 
      caption = "Sparsity Summary", col.names = "No. Missing Values")

## find the top10
kable(head(aifc[order(abs(aifc$aifc), decreasing = TRUE), ], 10), 
      caption = "Top10 Genes by Fold Change (Knock-out vs. WT)")



```

## 2. Obtain the p-values from a two sided t-test for differential expression. How many genes are significant at the 0.01 level? List the top 10 genes that have the largest t-statistics and their corresponding p-value. 
There are 75 genes are significant at the 0.01 level, by gene-level individual t-test.  

```{r ttest2}
# carry out individual t-tests

indi <- lapply(1:6384, function(row){
  test = t.test(ai[row, 2:9], ai[row, 10:17], alternative = "two.sided")
  test.sum = c(ai[row, 1] , test$p.value, test$statistic)
  test.sum
})

indi_t <- data.frame(matrix(unlist(indi), ncol = 3, byrow = TRUE))
colnames(indi_t) <- c("genenames", "pvalue", "tstatistic")
indi_t <- indi_t %>%mutate(pvalue = as.numeric(pvalue), 
                           tstatistic = as.numeric(tstatistic))  
indi_p <- indi_t %>% filter(pvalue <= 0.01)
dim(indi_p)

kable(head(indi_p[order(abs(indi_p$tstatistic), decreasing = TRUE), ], 10), 
              caption = "Top10 Genes by gene-level t-test")

                            

```


## c. Calculate the **modified** t-statistic and corresponding p-value using the samr package in R used in Homework2. How many genes are significant at the 0.01 level? List the top 10 genes that have the largest **penalized** t-statistics.

```{r samr}


```


## d. 

```{r limma}



```

# P-values and Multiple Testing 
##

```{r permutation}
```

##
```{r adjust}
# the original t-test 1b
n = nrow(ai)
head(indi_t, 3)

## the Bonferroni is n*p
indi_adjust <- indi_t %>% mutate(pBonferroni_FWER = pmin(1, n*pvalue),
                                 psidak_FWER = (1 - (1- pvalue)^n),
                                 pholm_FWER = ,
                                 pBH_FDR = ,
                                 pBY_FDR)

## check adjusted p-value 
head(indi_adjust, 3)
indi_adjust[540, ]

p <- c(0.01, 0.001, 0.002, 0.05)
o <- order(p)
ro <- order(o)
p
o
ro


```






