---
title: "R Markdown Tutorial"
author: "Mikaela Miller"
date: "Sunday, January 04, 2015"
output: word_document
---
# Introduction 

This document serves as a short introduction to R Markdown.  R Markdown can be used to generate reports that adhere to reproducible research initiatives, capturing computer code, datasets, methods, protocols and results in a single document. The following highlights many of the basic commands.  

## Resources

The links below are helpful references to get started in R Markdown:

  * http://rmarkdown.rstudio.com/authoring_basics.html
  * https://www.youtube.com/watch?v=YcJb1HBc-1Q

# Using Rmarkdown

There are two basic modes for R Markdown: Pandoc Markdown and R code chunks. Pandoc Markdown is where you will write the body of the report.  

Remember, always save your R Markdown file.  To generate the report, click on the "Knit ____" icon in the menu bar of the Rmd window.

## Pandoc

You can control different aspects of the text in Pandoc Markdown (size, italic, bullet points, etc.). Below are some examples. 

Emphasis

*italic*   **bold**

_italic_   __bold__

Headers

# Header 1

## Header 2

### Header 3

Lists

Unordered List:

* Item 1
* Item 2
    + Item 2a
    + Item 2b
    
Ordered List:

1. Item 1
2. Item 2
3. Item 3
    + Item 3a
    + Item 3b



## R code chunks
The execution of R code in Markdown is accomplished with R code chunks. You can embed the code and make it visible or hide it (which is useful when the code generates a table or graph).  Names can be given to each code chunk for navigation. Results can be cached to speed up processing time.

Generally, I like to make sure the code is working correctly by making an R Script in a separate file.  I then copy and paste the code chunks into the R Markdown file.

The most useful and frequently used commands:

* message=F (will hide any messages that are produced when the code is executed)
* warning=F (will hide any warnings that are produced when the code is executed)
* cache=T (caches results to speed up processing)
* echo=F (hides the code chunk, in the case you are creating a graphic)
* results="asis" (what I have found to be the best option for tables and graphics)



# Example Analysis Workflow

The remainder of the document runs through an example workflow. 

```{r qplot, fig.width=4, fig.height=3, message=FALSE}
# quick summary and plot
library(ggplot2)
summary(cars)
qplot( speed, dist, data=cars)
```


## Adding images to the R Markdown report

To add an immage from a file, you can use the following code and packages.  The image file must be in the working directory location. Replace the file location below with the location of an image you want to add. 

```{r, image, echo=F, eval=T, message=F, cache=F}
library(png)
library(grid)
img1 <- readPNG("RStudio-Ball.png")
grid.raster(img1)
```

#### References

You can cite R packages with the following code:

```{r, citations, echo=F, message=F}
citation("png")
```


#### Session Information

The session information contains the versions of R and all packages used in the analysis contained in your R Markdown file.

```{r, results='show', echo=F}
MA.session = sessionInfo(package = NULL)
print(MA.session)
```


